---
title: "7: Mail merge"
format: gfm
---

## Turning a spreadsheet of potential sources into customized emails

If you keep well-organized spreadsheets of potential sources' email addresses -- or if there are public directories of contact emails for potential sources that you _could_ turn into a spreadsheet -- [mail merge](https://support.google.com/mail/answer/12921167?hl=en) can help you to avoid copying and pasting versions of the same email to dozens of people. Instead, you can send as many customized emails as you need to send with a single click. 

If you're already working in R -- or want more control and reproducibility in your mail merge work flow -- this demo will show you how to mail merge using R code.

> Note: This post is adapted from my [sporadically-updated code blog](https://dipierro.github.io/updates/mail-merge/). And please mail merge responsibly, avoiding spammy emails that will discourage rather than encourage people to speak with you.

## Configuring Gmail

* Set up [two-factor authentication](https://support.google.com/accounts/answer/185839?hl=en&co=GENIE.Platform%3DDesktop). 

* [Create an app password](https://support.google.com/mail/answer/185833?hl=en). 

* Stash your password in an .Renviron file with the environmental variable `GMAIL_APP_PW` and `usethis::edit_r_environ()`.

## Formatting emails

Next, create an email template and use string substitution to customize each message. I've created a little toy data for this purpose, but you can read in a spreadsheet instead.

```{r}
#| message: false
#| warning: false
# Libraries
library(purrr)
library(curl)
library(stringr)

# Parameters
pw <- Sys.getenv("GMAIL_APP_PW") 
sender <- "adipierro@edsource.org"

# Code

## Recipients
recipients <- 
  tibble::tribble(
    ~first_name, ~last_name,	~email,	~employer,
    "Amy",	"DiPierro",	"adipierro@edsource.org",	"EdSource",
    "Mya", "DePriori",	"adipierro@edsource.org",	"CodeUser",
  )

## Write an email template
template <- "
From: 'Amy DiPierro' <adipierro@edsource.org>
To: '{first_name} {last_name}' <{email}>
Subject: Hello, {first_name}!

Dear {first_name},

How's it going at {employer}?

Cheers,
Amy
"

# Use string formatting to customize each email.
message <- recipients %>% str_glue_data(template)

# Extract list of recipients' emails
emails <- recipients %>% dplyr::pull(email)
```



Here's an example of how the email template looks after string formatting:

```{r}
#| message: false
#| warning: false
#| echo: false
message[2]
```

## Sending emails

Finally, just use the `send_mail` function from R's `curl` library  to email each customized message to the right recipient. The code below loops through two lists of customized messages and email addresses using the `purrr` function `map2`.

```{r}
#| message: false
#| warning: false
#| results: false
# Send all emails
map2(
  .x = message,
  .y = emails,
  .f = ~ send_mail(
    mail_from = sender,
    mail_rcpt = .y,
    message = .x,
    smtp_server = "smtps://smtp.gmail.com",
    username = sender,
    password  = pw,
    verbose = FALSE
  )
)
```

Within a few moments, your test emails should arrive in your email.
