---
title: "8: Query larger-than-memory databases"
format: gfm
---

## When your data is too big for memory

When I was first learning R a few years ago, I was fascinated by the [Federal Judicial Center's Integrated Database](https://www.fjc.gov/research/idb) (IDB), which provides case-related data about federal civil, criminal, bankruptcy and appellate court cases dating from 1970 to the present. Basically, it keeps tabs on every case in PACER.[^1]

There was one problem: Some of these files are pretty big. They're not _enormous_, but they're large enough that they can be slow to load and clunky to analyze in R. They didn't easily fit into memory on my laptop at the time, so it would freeze and RStudio would crash.

Typically, for a larger database of this kind, it will probably make sense to query the data using SQL. But depending on your comfort level with SQL, and the rest of your workflow, there are a growing number of R packages that make it easier to handle large databases without leaving R. 

Let's talk about two of those packages: `dbplyr` and `duckplyr`.

## Large-ish data with dplyr: The old way

Suppose you want to analyze all federal civil court cases that were filed, terminated or pending between the late 1980s and the present. That data is available from the Federal Judicial Center [here](https://www.fjc.gov/research/idb/civil-cases-filed-terminated-and-pending-sy-1988-present).

If you were working with `dplyr` only, you might do something like this.

> For the purpose of this demo, I've created a sample file that only contains the first 100 rows of FJC's most-recent civil data file. You can find the original data [here](https://www.fjc.gov/sites/default/files/idb/textfiles/cv88on.zip). You can read the rendered version of this fall to see accurate runtime estimates for analyzing the complete file in 08_databases.md.

```{r}
## Run this code during demo ##
# url_fjc <- "https://raw.githubusercontent.com/DiPierro/nicar26-10-things-R/refs/heads/main/sample-data/sample_fjc.csv"
# fjc_old <- readr::read_csv(url_fjc)
```

Here's the full dataset.

```{r}
# Do not run during demo ##
# This code actually downloads and reads in FJC data ##

# Set parameter for link to compressed file
zip_fjc <- "https://www.fjc.gov/sites/default/files/idb/textfiles/cv88on.zip"

# Create temporary file and directory to avoid a large download
temp_zip <- tempfile()
temp_dir <- tempfile()
dir.create(temp_dir)

# Download the file
download.file(zip_fjc, temp_zip, mode = "wb")
unzip(zipfile = temp_zip, exdir = temp_dir)
file_path <- file.path(temp_dir, "cv88on.txt")
file_details <- file.info(file_path)

# Read the file using readr::read_tsv
start_readr <- Sys.time()
fjc_old <- readr::read_tsv(file_path, col_names = TRUE, col_types = "c")
end_readr <- Sys.time()

print(end_readr - start_readr)
```


Here's the problem: The file is `r scales::number_bytes(x = file_details$size, symbol = "GiB")` and has `r scales::comma(nrow(fjc_old))` rows. Just reading it into R will take a long time.

Running even simple `dplyr` operations will also be a little sluggish.

```{r}
library(dplyr)

start_analyze <- Sys.time()

fjc_old |> 
  # Remove cases with a dummy termination date
  filter(TERMDATE != "01/01/1900") |> 
  # Add a column computing how long cases are pending
  mutate(time_pending = lubridate::mdy(TERMDATE) - lubridate::mdy(FILEDATE)) |> 
  # Summarize the average time it takes to resolve a lawsuit 
  # for each nature of suit (NOS) in each circuit court
  summarize(
    .by = c(NOS, CIRCUIT),
    mean_time_pending = mean(time_pending, na.rm = TRUE)
  ) |> 
  arrange(desc(mean_time_pending))
  
end_analyze <- Sys.time()
```

Here's how long that took to compute.

```{r}
print(end_analyze - start_analyze)
```

## Hello, dbplyr

`dbplyr` is a handy R package that could be useful for situations like this one -- and especially if your data is an order of magnitude larger. To quote its [documentation](https://dbplyr.tidyverse.org/):

> The goal of dbplyr is to automatically generate SQL for you so that you’re not forced to use it.

Its developers suggest that it's best for when:

* Your data is already in a database.

* Your data is too big to fit in memory.

We can give it a whirl with our FJC file like so. This is a little complex. What we're doing is that we're using two other R packages to connect to a database from R. The first package is called `DBI`. The second is called `RSQLite` and it's one of [several R packages](https://dbplyr.tidyverse.org/articles/dbplyr.html#getting-started) that you can use to connect to a database backend such as MySQL, Postgres or Google's BigQuery. 

In our case, we're creating a temporary SQLite database with no data in it.

```{r}
# Run once
# install.packages("RSQLite")

library(dbplyr)
con <- DBI::dbConnect(RSQLite::SQLite(), dbname = ":memory:")
```

Next, we're going to copy the FJC data to our empty database. We'll also create a reference to it by using `tbl()`.

```{r}
copy_to(
  dest = con, 
  df = 
    fjc_old |> 
    # Add a column computing how long cases are pending
    # dbplyr will struggle to find the SQL equivalent of this code
    mutate(time_pending = lubridate::mdy(TERMDATE) - lubridate::mdy(FILEDATE)), 
  name = "fjc_db",
  temporary = FALSE, 
  indexes = list("CIRCUIT", "NOS"),
  overwrite = TRUE
)

fjc_db <- tbl(con, "fjc_db")
```

Now, we can run the same code as we did before.

```{r}
start_dbplyr <- Sys.time()
fjc_db |> 
  # Remove cases with a dummy termination date
  filter(TERMDATE != "01/01/1900") |> 
  # Summarize the average time it takes to resolve a lawsuit 
  # for each nature of suit (NOS) in each circuit court
  summarize(
    .by = c(NOS, CIRCUIT),
    mean_time_pending = mean(time_pending, na.rm = TRUE)
  ) |> 
  arrange(desc(mean_time_pending))
end_dbplyr <- Sys.time()
```

I have to admit that the FJC database is a little on the small side for `dbplyr` to be super useful. But as you work with larger data -- or tap into databases that are already being kept in BigQuery or some other database -- `dbplyr` is a useful tool to have in your kit. Here's how long the query took to run:

```{r}
end_dbplyr - start_dbplyr
```

## Hello, duckplyr

The R package `duckplyr` combines the ease of writing `dplyr` commands like `filter()` and `summarize()` with the speed of the database management system [DuckDB](https://duckdb.org/).

Similar to `dbplyr` his magic sentence from the [documentation](https://duckplyr.tidyverse.org/) explains why this is so cool:

> The duckplyr package will run all of your existing dplyr code with identical results, using DuckDB where possible to compute the results faster. In addition, you can analyze larger-than-memory datasets straight from files on your disk or from the web.

Firing up `duckplyr` is as simple as...

```{r}
# Installation - Run once if on personal laptop
# install.packages("duckplyr")

# Load libraries
library(duckplyr)
library(dplyr)
conflicted::conflict_prefer("filter", "dplyr")
```

If you want to ingest a large file without downloading it, you can use this workflow:

```{r}
# DuckDB extension that makes it possible to query files in R without downloading them first
db_exec("INSTALL httpfs")
db_exec("LOAD httpfs")

# Another DuckDB extension that makes it possible to ingest zip files
# You can see the source code here: https://github.com/isaacbrodsky/duckdb-zipfs
db_exec("INSTALL zipfs FROM community")
db_exec("LOAD zipfs")

# Set a new parameter with the full zip file address
zip_fjc <- "zip://https://www.fjc.gov/sites/default/files/idb/textfiles/cv88on.zip/cv88on.txt"

# Read in FJC database using DuckDB
start_duck_read <- Sys.time()
fjc <- 
  read_csv_duckdb(
    file_path, 
    prudence = "thrifty", # We'll talk about this step in a moment
  )
end_duck_read <- Sys.time()
print(end_duck_read - start_duck_read)
```

Great. That was fast and easy.

```{r}
start_duck_analyze <- Sys.time()

fjc |> 
  # Filter out any cases in which a case was 'terminated' before it was filed
  # This typically indicates there is a placeholder termination date; the case might be open
  filter(TERMDATE > FILEDATE) |> 
  # Add a column computing how long cases are pending
  mutate(time_pending = TERMDATE - FILEDATE) |> 
  # Summarize the average time it takes to resolve a lawsuit 
  # for each nature of suit (NOS) in each circuit court
  summarize(
    .by = c(NOS, CIRCUIT),
    mean_time_pending = mean(time_pending, na.rm = TRUE)
  ) |> 
  arrange(desc(mean_time_pending))
  
end_duck_analyze <- Sys.time()
```

Now, our computation time should be a little faster, too.

```{r}
print(end_duck_analyze - start_duck_analyze)
```

## Working with 'prudence' in duckplyr

The joy of `duckplyr` is that you can control when to _materialize_ data, which helps us to avoid using too much memory by accident. The package lets users choose from three levels of _prudence_:

* *lavish* (careless about resources): Runs the computation with no questions asked. The FJC data, with just over 1 million rows and 46 columns, could be just slightly too big for this setting. 

* *stingy* (avoid spending at all cost): Always throws an error when accessing a column or printing the number of rows. You can override this by calling the function `collect()` before using any `dplyr` verbs.

* *thrifty* (use resources wisely): A compromise. Runs the computation if the data is small enough.

For anything that you can't do with `ducklyr`, the package will fallback to `dplyr` -- but only if a dataframe is  lavish or thrifty.

As an example, the following code will fail:

```{r}
# This code throws an error
# fjc |> nrow()
```

But we can override the _thrifty_ setting by using `collect()`.

```{r}
fjc |> 
  # Select first column to avoid read errors in the CSV
  select(CIRCUIT) |> 
  # Use collect() to override thrifty prudence setting
  collect() |> 
  nrow()
```

## More to explore

* ["R for Data Science" - Chapter 21, Databases](https://r4ds.hadley.nz/databases.html)

* ["Working with databases and SQL in RStudio"](https://posit.co/blog/working-with-databases-and-sql-in-rstudio/)

[^1]: If you want to use the IDB in your own work, I highly recommend reading this post from the [Free Law Project](https://free.law/idb-facts/) and this one from [SCALES OKN](https://livingreports.scales-okn.org/#/idbCrosswalkReport).