---
title: "10: Fetch YouTube data"
format: gfm
---

## Downloading comments and metadata from YouTube

By some [recent estimates](https://www.bbc.com/future/article/20250213-youtube-at-20-a-computer-that-drunk-dials-online-videos-reveals-statistics-that-google-doesnt-want-you-to-know), YouTube reaches almost one-third of people on earth.

Journalists and academic researchers are among those investigating the degree to which YouTube's recommendation algorithm may [have a radicalizing effect](https://www.asc.upenn.edu/news-events/news/youtube-algorithm-isnt-radicalizing-people) or [contribute to the spread of disinformation](https://www.npr.org/2021/04/13/986678544/exploring-youtube-and-the-spread-of-disinformation). And YouTube's size alone means that individual YouTubers can be globally influential.

In short, YouTube can be a rich data source. An R package called [`tuber`](https://github.com/gojiplus/tuber) makes it easier to tap into that data using R.

## Getting started

> Warning: This process can be tricky and difficult to debug. Feel free to sit back and watch the demo for now. 

1. Go to the [Google APIs dashboard](https://console.developers.google.com/apis/dashboard) and enable APIs and services if you haven't already.

2. Create a new project. Call it 'tuber' or something else descriptive.

3. Use the search bar to find the YouTube Data API v3 and click 'Enable'.

4. Go to 'Create credentials' > 'OAuth client ID' and select 'Desktop application' as the application type. Name the application 'tuber-rstudio' or something similar so you know what it is later.

5. Go to 'Audience' in the side panel and add your own gmail address as a test user.

Make sure to save your app ID and password. You can then stash them in your .Renviron the usual way.

```{r}
# Set YT_ID and YT_PASSWORD variables, then restart R
usethis::edit_r_environ(scope = "project")
```

Then, you can authenticate like this.

```{r}
# Libraries
library(tuber)
library(dplyr)
library(scales)

# Here's how to authenticate
app_id <- Sys.getenv("YT_ID")
app_password <- Sys.getenv("YT_PASSWORD")
yt_oauth(app_id, app_password, token = "")
```

## Hello, tuber

By many metrics, [MrBeast](https://en.wikipedia.org/wiki/MrBeast) ranks among the most popular YouTube channels of all time. Among the channel's most popular videos as of this writing is November 2021's ["$456,000 Squid Game In Real Life!
"](https://www.youtube.com/watch?v=0e3GPea1Tyg) has more than 900 million views to date. 

Using `tuber`, we can start to peruse its impressive statistics with a few lines of code:

```{r}
video_id <- "0e3GPea1Tyg"

# See how many times a video was viewed and liked
video_stats <- get_stats(video_id = video_id)

cat(
  "The video has", 
  comma(as.double(video_stats$viewCount)),
  "views,",
  comma(as.double(video_stats$likeCount)),
  "likes and",
  comma(as.double(video_stats$commentCount)),
  "comments."
)
```

We can also dig into other details, including tags and more:

```{r}
# See tags, video description and more
details <- get_video_details(video_id)

# Grab all of the video tags
details$items[[1]]$snippet$tags |>
  unlist() |> 
  as_data_frame()
```

Here's where to find the video description:

```{r}
cat(details$items[[1]]$snippet$description) 
```


## More to explore

Unfortunately, there are a lot of useful things you can't use `tuber` to do. If working with YouTube videos and data becomes a part of your reporting, there is a lot of software outside of R that you can explore. For example:

* You can't lift captions off a YouTube video. 

* You also can't download YouTube videos. For that, you might try [`yt-dlp`](https://github.com/yt-dlp/yt-dlp).

* Also, Google long ago ditched a parameter [used to retrieve related videos](https://developers.google.com/youtube/v3/revision_history#release_notes_06_12_2023) and is [suing a company](https://blog.google/innovation-and-ai/technology/safety-security/serpapi-lawsuit/) that provides a workaround.




